% ************************** Thesis Acknowledgements **************************

\begin{acknowledgements}      

When I began this thesis, I could never have imagined how much I would learn and grow. It would not have been possible without the following people.

\noindent
First, my utmost appreciation goes to my supervisor Zamin Iqbal. Zam, your insatiable curiosity and razor-sharp intellect are a true wonder. I could not imagine a better guide on this journey.

\noindent
I am forever grateful to Rachel, Leandro, and Robyn for their unrelenting effort and instruction. I will always cherish the enlightening whiteboard sessions, cafeteria discussions, and convoluted Slack and GitHub messages.  

\noindent
My heartfelt thanks to all members of the Iqbal group, past and present. Kerri, Grace, Martin, Brice, Leah and everyone else; you shaped the scientist I have become.

\noindent
To Simon Grandjean Lapierre, whose clinical perspective helped keep me grounded in the real world. I would also like to acknowledge the many insightful conversations with other collaborators: Marie Sylvianne Rabodoarivelo, Anastasia Koch, Niaina Rakotosamimanana, Anzaan Dippenaar, Helen Cox, and Tim Peto.

\noindent
Thank you to my thesis advisory committee members: John Marioni, Simon Frost, Georg Zeller, and Estée Török. Your guidance and encouragement were vital.

\noindent
Lastly, I would like to acknowledge EMBL for their generous financial support and providing such a supportive environment to undertake a PhD in.

\end{acknowledgements}
